---
title: "Data"
date: "`r format(Sys.Date())`."
output:
  html_document:
    theme: readable
    toc: yes
    number_sections: true
    toc_depth: 4
    toc_float: 
      collapsed: true
    code_download: yes
    df_print: paged
---

<style type="text/css">
body, td{ font-size: 14px; }
code.r{ font-size: 12px; }
pre{ font-size: 12px }
h1.title {
  font-size: 24px;
  color: Black;
}
h1 { /* Header 1 */
  font-size: 20px;
  color: Black;
}
h2 { /* Header 2 */
  font-size: 16px;
  color: Black;
}
h3 { /* Header 3 */
    font-size: 14px;
  color: Black;
}
h4 { /* Header 4 */
    font-size: 12px;
  color: Black;
}
#TOC {
  color: DarkBlue;
  font-size: 16px;
}
</style>

```{r include=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = FALSE, comment="", warning = FALSE,
                      message=FALSE, fig.align = "center")
```

# Definir path local

```{r}
path <- here::here()
setwd(path)
```

# Paquetes

```{r}
library(ggplot2)
library(spatstat)
library(fpp3)
library(leaflet)
library(sf)
#perfil del continente
library(rnaturalearth)
```


# Data

Sismic events of ['https://earthquake.usgs.gov/'] from 01-01-1901 to 01-01-2025

```{r cars}
data <- readRDS('input/sismicevents.rds')
data$time <- as.POSIXct(data$time, tz = "UTC")
```

```{r}
data |> st_drop_geometry() |> 
  select(time, magnitude, depth) |> head()
```
```{r}
data |> st_drop_geometry() |> 
  select(time, magnitude, depth) |> tail()
```

# Time vs Magnitude

You can also embed plots, for example:

```{r pressure, echo=FALSE}
ggplot(data, aes(time, magnitude)) + 
  geom_point(size = 1, alpha = 0.25) +
  scale_x_datetime(
    date_breaks = "10 years",
    date_labels = "%Y"
  ) + geom_vline(xintercept =  as.POSIXct("1980-01-01 00:00:00"),
                 linetype="dotted", 
                color = "red", size=0.5)+
  coord_cartesian(
    xlim = c(
      as.POSIXct("1902-01-01", tz = "UTC"),
      as.POSIXct("2025-07-30", tz = "UTC")
    ), 
    ylim = c(0,10)
  ) +
  theme_minimal()
```

## Magnitude

```{r}
table(data$magType)
```
Body-Wave Magnitude — mb
Surface-Wave Magnitude — ms
Moment Magnitude — mw

Variants of Mw
mwb → moment magnitude (body-wave inversion)
mwc → centroid moment tensor magnitude
mww → W-phase inversion
mwr → regional moment magnitude

```{r}
x <- data |> filter(magType == "mw") 
table(year(x$time))
```

```{r}
ggplot(data, aes(x = time, y = magnitude)) +
  geom_point(size = 0.1) +
  ggtitle("Horus catalogue magnitude time series")
```

```{r}
mw_family <- c("mw", "mwb", "mwc", "mwr", "mww")
dtmag <- data |> filter(magType %in% mw_family) |>
            filter(magnitude > 5)
```


# Locations

```{r}
continent <- ne_countries(continent = "South America", returnclass = "sf")

paises <- c("COL", "ECU", "PER", "CHL")
cont <- continent[continent$adm0_a3 %in% paises, ] |> st_make_valid()
cont_u <- st_union(cont) |> st_make_valid()

# Natural Earth physical coastline (LINESTRING)
coast <- ne_download(scale = 50, type = "coastline", category = "physical",
                     returnclass = "sf") |> st_make_valid()

# Match CRS
coast <- st_transform(coast, st_crs(cont_u))

# Keep only coastline segments that fall on/near your selected countries
perfil <- st_intersection(coast, cont_u)

plot(st_geometry(perfil), col = "blue")
```

```{r}
library(sf)
library(lwgeom)  # para st_make_valid (muy recomendable)

sf_use_s2(TRUE)

ancho_km <- 400
buffer_costa <- st_union(st_buffer(perfil, dist = ancho_km * 1000))
buffer_costa <- st_make_valid(buffer_costa)  # arregla geometrías inválidas

# Caja Falklands en 4326
falklands_box <- st_as_sfc(
  st_bbox(c(xmin = -64, xmax = -55, ymin = -62, ymax = -47), crs = st_crs(4326))
)

# Proyectar a un CRS planar para hacer difference de forma estable
buffer_3857 <- st_transform(buffer_costa, 3857) |> st_make_valid()
falk_3857   <- st_transform(falklands_box, 3857) |> st_make_valid()

buffer_sin_falk_3857 <- st_difference(buffer_3857, falk_3857) |> st_make_valid()
buffer <- st_transform(buffer_sin_falk_3857, 4326)

plot(st_geometry(buffer))
```






```{r}
# Unificar franja para hacer el filtro más rápido

dt <- dtmag |> filter(year(time) > 1980 )
sismos <- dt[st_within(dt, buffer, sparse = FALSE), ]

plot(st_geometry(sismos))
```



```{r}
pal <- colorNumeric(
  palette = "viridis",
  domain  = data$magnitude,
  reverse = TRUE
)

sismos |> 
  # filter(magnitude > 2.5) |>
  leaflet() |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude,
    radius = 0.5,
    color = ~pal(magnitude),
    opacity = 0.7,
    fillOpacity = 0.6
  ) |>
  addLegend(
    pal = pal,
    values = ~magnitude,
    title = "Magnitude"
  )

```

```{r}
saveRDS(sismos, "input/sismos.rds")
```


# Exploratory

```{r}
dim(sismos)
```

```{r}
library(sf)
library(ggplot2)
library(patchwork)

# Check CRS
st_crs(perfil)
st_crs(buffer)
st_crs(sismos)

# If needed, transform everything to the same CRS (e.g., WGS84)
perfil  <- st_transform(perfil, 4326)
buffer  <- st_transform(st_as_sf(buffer), 4326)
sismos  <- st_transform(sismos, 4326)
```

## Space

```{r}
map_buffer <- ggplot() +
  geom_sf(data = buffer, fill = "lightgreen", alpha = 0.4, color = NA) +
  geom_sf(data = perfil, color = "brown", linewidth = 0.6) +
  labs(title = "(a)")+#Pacific Coastline and 400 km Buffer") +
  theme_minimal() +
  theme(
    panel.grid.major = element_line(color = "grey85"),
    plot.title = element_text(face = "bold")
  )
map_buffer
```

```{r}
map_points <- ggplot() +
  geom_sf(data = sismos, color = "darkgreen", size = 0.3, alpha = 0.5) +
  labs(title = "(b)")+#Seismic Events (1980–2025)") +
  theme_minimal() +
  theme(
    panel.grid.major = element_line(color = "grey85"),
    plot.title = element_text(face = "bold")
  )
map_points
```

```{r}
bbox <- st_bbox(buffer)

map_buffer <- map_buffer + coord_sf(
  xlim = c(bbox["xmin"], bbox["xmax"]),
  ylim = c(bbox["ymin"], bbox["ymax"]),
  expand = FALSE
) +
  scale_x_continuous(
    breaks = seq(-90, -60, by = 10)
  )

map_points <- map_points + coord_sf(
  xlim = c(bbox["xmin"], bbox["xmax"]),
  ylim = c(bbox["ymin"], bbox["ymax"]),
  expand = FALSE
) +
  scale_x_continuous(
    breaks = seq(-90, -60, by = 10)
  )
```

```{r}
map_buffer + map_points
```


```{r}
combined_map <- map_buffer + map_points
ggsave(
  filename = "figure_study_region.png",
  plot = combined_map,
  width = 12,
  height = 6,
  dpi = 300
)
```


## Time

```{r}
library(dplyr)
library(lubridate)
library(ggplot2)
library(fpp3)
```

```{r, fig.width= 6, fig.height= 3}
sismos$year  <- year(sismos$time)
sismos$month <- floor_date(sismos$time, "month")

# Yearly counts
year_counts <- sismos %>%
  count(year)

gtime <- ggplot(year_counts, aes(year, n)) +
  geom_line() +
  geom_point() +
  labs(title = "(c) Yearly Earthquake Counts",
       x = "Year", y = "Number of Events") +
  theme_minimal()
gtime
```

```{r}
ggsave(
  filename = "figure_time.png",
  plot = gtime,
  width = 6,
  height = 3,
  dpi = 300
)
```

```{r}
combined_map <- map_buffer + map_points
ggsave(
  filename = "figure_combined.png",
  plot = map_buffer + map_points + gtime,
  # width = 12,
  height = 6,
  dpi = 300
)
```

## Magnitud

```{r}
table(sismos$magType)
```



```{r}
map_magnitude <- ggplot(sismos, 
                        aes(x = time, y = magnitude)) +
  geom_point(size = 0.1) +
  ggtitle("(c) Horus catalogue magnitude time series")+
  theme_minimal()
map_magnitude
```

```{r}
combined_map <- map_buffer + map_points
ggsave(
  filename = "figure_combined.png",
  plot = map_buffer + map_points + map_magnitude,
  width = 12,
  height = 6,
  dpi = 300
)
```


# Gutenberg–Richter Analysis (Critical for ETAS)

```{r}
ggplot(sismos, aes(magnitude)) +
  geom_histogram(binwidth = 0.1,
                 fill = "grey70",
                 color = "black") +
  theme_minimal() +
  labs(title = expression("Histogram of Moment Magnitudes (" * M >= 5 * ")"),
       x = "Moment Magnitude (Mw)",
       y = "Frequency")
```
```{r}
M0 <- 5
mag_sorted <- sort(sismos$magnitude)

unique_mag <- sort(unique(mag_sorted))

Ncum <- sapply(unique_mag, function(m) {
  sum(mag_sorted >= m)
})

gr_df <- data.frame(
  m = unique_mag,
  logN = log10(Ncum)
)
```



```{r}
ggplot(gr_df, aes(m, logN)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  theme_minimal() +
  labs(title = "Gutenberg–Richter Plot",
       x = "Magnitude (Mw)",
       y = expression(log[10]~N(M >= m)))
```

```{r}
gr_sub <- gr_df %>% filter(m >= 5)

fit <- lm(logN ~ m, data = gr_sub)
summary(fit)
```

By choosing: $M_0=5$
* Linear Gutenberg–Richter behavior
* Physically plausible b-value
* High explanatory power
* No obvious magnitude-scale instability
* This validates your threshold choice.

## Temporal stability of b-value

Roughly stable mean → stationary background rate.
```{r}
sismos$year <- year(sismos$time)

year_counts <- sismos%>% count(year)

ggplot(year_counts, aes(year, n)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = 
         expression("Yearly Counts (" * M >= 5 * ")"),
       x = "Year",
       y = "Number of Events")
```

β1 significant → background rate may be nonstationary.
```{r}
trend_fit <- lm(n ~ year, data = year_counts)
summary(trend_fit)
```

In seismicity, expect vmr >> 1 due to aftershocks.
```{r}
vmr <- var(year_counts$n) / mean(year_counts$n)
vmr
```

Linear growth → stationary rate.
```{r}
year_counts <- year_counts %>%
  arrange(year) %>%
  mutate(cumN = cumsum(n))

ggplot(year_counts, aes(year, cumN)) +
  geom_line() +
  theme_minimal() +
  labs(title = 
         expression("Cumulative Counts (" * M >= 5 * ")"),
       x = "Year",
       y = "Cumulative Events")
```
The LOESS curve shows underlying drift ignoring bursts.

```{r}
ggplot(year_counts, aes(year, n)) +
  geom_line(alpha = 0.5) +
  geom_smooth(method = "loess", span = 0.3, se = FALSE, color = "red") +
  theme_minimal() +
  labs(title = 
  expression("Smoothed Yearly Counts (" * M >= 5 * ")"))
```

```{r}
glm_fit <- glm(n ~ year, 
               family = poisson(link = "log"), 
               data = year_counts)

summary(glm_fit)
```


<!-- What This Does NOT Guarantee -->
<!-- We still need to check: -->
<!-- Temporal stability of b-value -->
<!-- Stability across decades -->
<!-- Stability across space (coastal segments) -->
<!-- Because if b changes over time, ETAS parameters may drift. -->



<!-- Check temporal stationarity of counts (M ≥ 5) -->
<!-- ```{r} -->
<!-- library(ggplot2) -->

<!-- ggplot(sismos, aes(magnitude, fill = magType)) + -->
<!--   geom_density(alpha = 0.4) + -->
<!--   theme_minimal() + -->
<!--   labs(title = "Magnitude Density by Mw-Type", -->
<!--        x = "Magnitude", y = "Density") -->
<!-- ``` -->
<!-- ```{r} -->
<!-- library(ggplot2) -->

<!-- ggplot(sismos, aes(magnitude, fill = magType)) + -->
<!--   geom_density(alpha = 0.4) + -->
<!--   theme_minimal() + -->
<!--   labs(title = "Magnitude Density by Mw-Type", -->
<!--        x = "Magnitude", y = "Density") -->
<!-- ``` -->


<!-- ```{r} -->
<!-- sismos <- sismos %>% -->
<!--   arrange(time) -->

<!-- tau <- diff(as.numeric(sismos$time))  # in minutes -->
<!-- ``` -->

<!-- ```{r} -->
<!-- ggplot(data.frame(tau = tau), aes(tau)) + -->
<!--   geom_histogram(bins = 100) + -->
<!--   scale_x_log10() + -->
<!--   labs(title = "Distribution of Inter-event Times (log scale)", -->
<!--        x = "Inter-event time (minutes)", y = "Frequency") + -->
<!--   theme_minimal() -->
<!-- ``` -->

<!-- ```{r} -->
<!-- sismos$cumN <- 1:nrow(sismos) -->

<!-- ggplot(sismos, aes(time, cumN)) + -->
<!--   geom_line() + -->
<!--   labs(title = "Cumulative Number of Events", -->
<!--        x = "Time", y = "Cumulative Count") + -->
<!--   theme_minimal() -->
<!-- ``` -->


<!-- # Other -->

<!-- ```{r} -->
<!-- # Add an integer id to each event -->
<!-- data$event_num <- seq.int(nrow(data)) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- ## This is just the EPSG equivalent of WGS84 -->
<!-- crs_wgs84 <- st_crs("EPSG:4326") -->
<!-- # Generate an sf version of the catalogue where the longitude and latitude are -->
<!-- # converted to a point object -->
<!-- df_cat.sf <- st_as_sf(data, -->
<!--   coords = c("longitude", "latitude"), -->
<!--   crs = crs_wgs84 -->
<!-- ) -->
<!-- ``` -->


<!-- ```{r} -->
<!-- hist( -->
<!--   data$magnitude -->
<!-- ) -->
<!-- ``` -->



<!-- ```{r} -->
<!-- # data$Time <- as.Date(data$time) -->

<!-- # print("Extract year") -->
<!-- # extract the year and convert to numeric format -->
<!-- data$Year <- as.numeric(format(data$time, "%Y")) -->

<!-- countEventsInYear <- data %>% -->
<!--   st_drop_geometry() %>% -->
<!--   group_by(Year) %>% -->
<!--   summarize(counts = n()) -->

<!-- plot( -->
<!--   countEventsInYear, -->
<!--   type = "l", -->
<!--   main = "Annual number of events for whole catalogue" -->
<!-- ) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- ggplot() + -->
<!--   geom_hex(data = data[data$magnitude >= 0, ],  -->
<!--            aes(x = longitude, y = latitude), bins = 50) + -->
<!--   scale_fill_continuous(type = "viridis") + -->
<!--   ggtitle("Density plot for M>3 events") + -->
<!--   theme_bw() -->
<!-- ``` -->



<!-- ```{r} -->
<!-- ggplot() + -->
<!--   geom_sf(data = data[data$magnitude > 0, ], size = 0.05) + -->
<!--   geom_sf(data = data[data$magnitude > 3, ],  -->
<!--           size = 0.5, color = "orange") + -->
<!--   geom_sf(data = data[data$magnitude > 6, ],  -->
<!--           size = 0.5, color = "red") + -->
<!--   ggtitle("Map of event locations") -->
<!-- ``` -->









