---
title: "Data"
date: "`r format(Sys.Date())`."
output:
  html_document:
    theme: readable
    toc: yes
    number_sections: true
    toc_depth: 4
    toc_float: 
      collapsed: true
    code_download: yes
    df_print: paged
---

<style type="text/css">
body, td{ font-size: 14px; }
code.r{ font-size: 12px; }
pre{ font-size: 12px }
h1.title {
  font-size: 24px;
  color: Black;
}
h1 { /* Header 1 */
  font-size: 20px;
  color: Black;
}
h2 { /* Header 2 */
  font-size: 16px;
  color: Black;
}
h3 { /* Header 3 */
    font-size: 14px;
  color: Black;
}
h4 { /* Header 4 */
    font-size: 12px;
  color: Black;
}
#TOC {
  color: DarkBlue;
  font-size: 16px;
}
</style>

```{r include=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = FALSE, comment="", warning = FALSE,
                      message=FALSE, fig.align = "center")
```

# Definir path local

```{r}
path <- here::here()
setwd(path)
```

# Paquetes

```{r}
library(ETAS.inlabru)
library(ggplot2)

library(rnaturalearth)
library(terra)
library(sf)
library(ggspatial)
library(rnaturalearthdata)
library(dplyr)
library(lubridate)

## This is just the EPSG equivalent of WGS84
crs_wgs84 <- st_crs("EPSG:4326")
```

```{r}
# install.packages("remotes")
# remotes::install_github("edinburgh-seismicity-hub/ETAS.inlabru")
```

```{r}
# library(ETAS.inlabru)
# library(ggplot2)

# Increase/decrease num.cores if you have more/fewer cores on your computer.
# future::multisession works on both Windows, MacOS, and Linux
num.cores <- 4
future::plan(future::multisession, workers = num.cores)
INLA::inla.setOption(num.threads = num.cores)
# To deactivate parallelism, run
#   future::plan(future::sequential)
#   INLA::inla.setOption(num.threads = 1)
```

# Function

```{r}
# set copula transformations list
link.f <- list(
  mu = \(x) gamma_t(x, 0.3, 0.6),
  K = \(x) unif_t(x, 0, 10),
  alpha = \(x) unif_t(x, 0, 10),
  c_ = \(x) unif_t(x, 0, 10),
  p = \(x) unif_t(x, 1, 10)
)

# set inverse copula transformations list
inv.link.f <- list(
  mu = \(x) inv_gamma_t(x, 0.3, 0.6),
  K = \(x) inv_unif_t(x, 0, 10),
  alpha = \(x) inv_unif_t(x, 0, 10),
  c_ = \(x) inv_unif_t(x, 0, 10),
  p = \(x) inv_unif_t(x, 1, 10)
)
```


# Data

Sismic events of ['https://earthquake.usgs.gov/'] 

```{r cars}
data <- readRDS('process_data/sismos.rds')
data$time <- as.POSIXct(data$time, tz = "UTC")
```

```{r}
# Add an integer id to each event
data$event_num <- seq.int(nrow(data))
```

```{r}
## This is just the EPSG equivalent of WGS84
crs_wgs84 <- st_crs("EPSG:4326")
# Generate an sf version of the catalogue where the longitude and latitude are
# converted to a point object
df_cat.sf <- st_as_sf(data,
  coords = c("longitude", "latitude"),
  crs = crs_wgs84
)
```

# Initialization

```{r}
# There may be some incorrectly registered data-times in the original data set,
# that as.POSIXct() can't convert, depending on the system.
# These should ideally be corrected, but for now, we just remove the rows that
# couldn't be converted.
dt <- na.omit(data)

# set up parameters for selection
start.date <- as.POSIXct("2008-01-01T00:00:00", 
                         format = "%Y-%m-%dT%H:%M:%OS")
end.date <- as.POSIXct("2010-01-01T00:00:00", 
                       format = "%Y-%m-%dT%H:%M:%OS")
# min.longitude <- -120
# max.longitude <- -32
# min.latitude <- -60
# max.latitude <- 33
M0 <- 2.5

# set up conditions for selection
aquila.sel <- (data$time >= start.date) &
  (data$time < end.date) &(data$magnitude >= M0)
  # (data$longitude >= min.longitude) &
  # (data$longitude <= max.longitude) &
  # (data$latitude >= min.latitude) &
  # (data$latitude <= max.latitude) &
  

# select
aquila <- data[aquila.sel, ]
```

```{r}
ggplot(aquila, aes(time, magnitude)) +
  geom_point() +
  theme_bw()
```

# Data preparation to model fitting

```{r}
# set up data.frame for model fitting
aquila.bru <- data.frame(
  ts = as.numeric(
    difftime(aquila$time, start.date, units = "days")
  ),
  magnitudes = aquila$magnitude,
  idx.p = seq_len(nrow(aquila))
)

# set up list of initial values
th.init <- list(
  th.mu = inv.link.f$mu(0.5),
  th.K = inv.link.f$K(0.1),
  th.alpha = inv.link.f$alpha(1),
  th.c = inv.link.f$c_(0.1),
  th.p = inv.link.f$p(1.1)
)
```

```{r}
# set up list of bru options
bru.opt.list <- list(
  bru_verbose = 3, # type of visual output
  bru_max_iter = 70, # maximum number of iterations
  # bru_method = list(max_step = 0.5),
  bru_initial = th.init # parameters' initial values
)
```

# Model fitting

```{r}
# set starting and time of the time interval used for model fitting. In this
# case, we use the interval covered by the data.
T1 <- 0
# Use max(..., na.rm = TRUE) if there may still be NAs here
T2 <- max(aquila.bru$ts) + 0.2
# fit the model
aquila.fit <- Temporal.ETAS(
  total.data = aquila.bru,
  M0 = M0,
  T1 = T1,
  T2 = T2,
  link.functions = link.f,
  coef.t. = 1,
  delta.t. = 0.1,
  N.max. = 5,
  bru.opt = bru.opt.list
)
```

# Create input list

```{r}
# create input list to explore model output
input_list <- list(
  model.fit = aquila.fit,
  link.functions = link.f
)
```

# Check marginal posterior distributions

```{r}
# get marginal posterior information
post.list <- get_posterior_param(input.list = input_list)

# plot marginal posteriors
post.list$post.plot
```

# Sample the joint posterior and make pair plot


```{r}
post.samp <- post_sampling(
  input.list = input_list,
  n.samp = 1000,
  max.batch = 1000,
  ncore = num.cores
)

head(post.samp)
```

```{r}
pair.plot <- post_pairs_plot(
  post.samp = post.samp,
  input.list = NULL,
  n.samp = NULL,
  max.batch = 1000
)
pair.plot$pair.plot
```

# Check posterior number of events

```{r}
# set additional elements of the list
input_list$T12 <- c(T1, T2)
input_list$M0 <- M0
input_list$catalog.bru <- aquila.bru
```

```{r}
N.post <- get_posterior_N(input.list = input_list)
N.post$post.plot
```

# Posterior of the triggering function and Omori law

```{r}
triggering_fun_plot(
  input.list = input_list,
  post.samp = post.samp,
  n.samp = NULL, magnitude = 4,
  t.end = 5, n.breaks = 100
)
```
```{r}
triggering_fun_plot_prior(
  input.list = input_list,
  magnitude = 4,
  n.samp = 1000,
  t.end = 10
)
```

```{r}
omori_plot_posterior(
  input.list = input_list,
  post.samp = post.samp,
  n.samp = NULL,
  t.end = 5
)
```
```{r}
omori_plot_prior(input.list = input_list, 
                 n.samp = 1000, t.end = 5)
```

# Generate synthetic catalogues from model

```{r}
# maximum likelihood estimator for beta
beta.p <- 1 / (mean(aquila.bru$magnitudes) - M0)
```

```{r}
# synth.cat.list <- generate_temporal_ETAS_synthetic(
#   theta = post.samp[1, ], # ETAS parameters
#   beta.p = beta.p, # magnitude distribution parameter
#   M0 = M0, # cutoff magnitude
#   T1 = T1, # starting time
#   T2 = T2, # end time
#   Ht = aquila.bru[which.max(aquila.bru$magnitudes), ] # known events
# )
# # merge into unique data.frame
# synth.cat.df <- do.call(rbind, synth.cat.list)
# # order events by time
# synth.cat.df <- synth.cat.df[order(synth.cat.df$ts), ]
# 
# ggplot(synth.cat.df, aes(ts, magnitudes, color = as.factor(gen))) +
#   geom_point(size = 0.5)
```

```{r}
# set.seed(2)
# n.cat <- 8
# # generate catalogues as list of lists
# multi.synth.cat.list <- lapply(seq_len(n.cat), \(x) {
#   generate_temporal_ETAS_synthetic(
#     theta = post.samp[x, ],
#     beta.p = beta.p,
#     M0 = M0,
#     T1 = T1,
#     T2 = T2,
#     Ht = aquila.bru[which.max(aquila.bru$magnitudes), ]
#   )
# })
# 
# # store catalogues as list of data.frames
# multi.synth.cat.list.df <- lapply(multi.synth.cat.list, \(x) do.call(rbind, x))
# # set catalogue identifier
# multi.synth.cat.list.df <- lapply(
#   seq_len(n.cat),
#   \(x) {
#     cbind(multi.synth.cat.list.df[[x]], cat.idx = x)
#   }
# )
# # merge catalogues in unique data.frame
# multi.synth.cat.df <- do.call(rbind, multi.synth.cat.list.df)
# 
# # we need to bing the synthetics with the observed catalogue for plotting
# cat.df.for.plotting <- rbind(
#   multi.synth.cat.df,
#   cbind(aquila.bru[, c("ts", "magnitudes")],
#     gen = NA,
#     cat.idx = "observed"
#   )
# )
# 
# # plot them
# ggplot(cat.df.for.plotting, aes(ts, magnitudes)) +
#   geom_point(size = 0.5) +
#   geom_point(
#     data = aquila.bru[which.max(aquila.bru$magnitudes), ],
#     mapping = aes(ts, magnitudes),
#     color = "red"
#   ) +
#   facet_wrap(facets = ~cat.idx)
```

# Forecasting


```{r}
# # express 1 minute in days
# min.in.days <- 1 / (24 * 60)
# # find time of the event with the greatest magnitude
# t.max.mag <- aquila.bru$ts[which.max(aquila.bru$magnitudes)]
# # set starting time of the forecasting period
# T1.fore <- t.max.mag + min.in.days
# # set forecast length
# fore.length <- 1
# # set end time of the forecasting period
# T2.fore <- T1.fore + fore.length
# # set known data
# Ht.fore <- aquila.bru[aquila.bru$ts < T1.fore, ]
# 
# # produce forecast
# daily.fore <- Temporal.ETAS.forecast(
#   post.samp = post.samp, # ETAS parameters posterior samples
#   n.cat = nrow(post.samp), # number of synthetic catalogues
#   beta.p = beta.p, # magnitude distribution parameter
#   M0 = M0, # cutoff magnitude
#   T1 = T1.fore, # forecast starting time
#   T2 = T2.fore, # forecast end time
#   Ht = Ht.fore, # known events
#   ncore = num.cores
# ) # number of cores
```


```{r}
# # find number of events per catalogue
# N.fore <- vapply(
#   seq_len(daily.fore$n.cat),
#   \(x) sum(daily.fore$fore.df$cat.idx == x), 0
# )
# # find number of observed events in the forecasting period
# N.obs <- sum(aquila.bru$ts >= T1.fore & aquila.bru$ts <= T2.fore)
# # plot the distribution
# ggplot() +
#   geom_histogram(aes(x = N.fore), binwidth = 1) 
#   # geom_vline(xintercept = N.obs) +
#   # xlim(100, 500)


```

